Working from home and digital divides: resilience during the pandemic

### 2020-12-30 paper/v1/broadband.speed.covid.Rmd
Added reference, figure dimensions, further edits to results, etc.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds
during the spring 2020 lockdown,
may reinforce or redress the spatial and social dimensions of digital divisions.
We employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of both occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era,
and broadband services and infrastructure must be fit for purpose.

@@

As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and interactions
between digital divisions geographically and socioeconomically.
By asking how resilient broadband speeds, and particularly upload speeds, are
experienced in different parts of the UK during a time of extreme demand,
we interrogate which places benefit from the greater economic resilience
digital technologies can offer.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand their structural and spatial development pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across clusters of UK local authorities and then assesses whether these clusters
replicate or repudiate other socio-economic and geographic patterns of economic
resilience.
We conclude with a discussion of the insights we have gained from our new
perspective on digital divisions.

@@

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies.
Rural areas, predictably, fare worse.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been investigated, in part because telecommuting has
not previously been a cause of concentrated demand and pressure on internet services.

@@

The extreme demand during the pandemic thus provides a new opportunity to understand
how internet infrastructure quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive 
video conferencing required in order to avoid the face-to-face contact that could 
increase the spread of infection.
We seek to answer how internet service resilience contributes to or reduces economic
resilience when the latter is dependent upon the capability to work from home.
We also aim to improve our understanding of the impact of first level digital
division on telecommuting, and whether this results in more fundamental third
level digital divisions than has previously been perceived.

@@

![Speed tests over time, 2019 \label{test2019}](figures/time.var.plot2019.png){height=40% width=75%}

@@

Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
clusters $1$, $13$, and $6$. However, mean upload speeds are much lower between
$9:00-10:59$ than between $19:00-20:59$ in cluster $8$, but slightly faster in the
morning than in the evening in cluster $12$.
Indeed, cluster $12$ is the only cluster to experience higher speeds in the morning
peak compared to the evening, and thus the only cluster where the temporal profile
of internet use is closer to what might have been expected pre-pandemic.
And yet even among the other clusters, the reliability of internet services
during the working day still varies considerably.
Interpreting this variation from the large spikes and dips shown on
Figures \ref{UpClusterL} and \ref{UpClusterS} is difficult,
but the statistics in Table \ref{up.cluster.descr} show that clusters $9$ and $11$
have the most reliable internet services.
The ratio of standard deviation to mean in both these clusters is lowest,
and speeds are only about 2% slower in the morning than the evening.
Mean speeds are also higher than in any other cluster, excluding cluster $2$,
where measures of reliability suggest poorer performance.

Thus, broadband services in clusters $9$ and $11$, 
home to over twelve million people, performed the best during the study period, 
in terms of both quality and reliability.
In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
with the lowest points occurring, on average, between $6:00-7:00$ on Monday morning, 
$14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday, not the workday peak. 
Furthermore, these slow periods offer better speeds than the average hourly
profile of cluster $13$. Finally, ignoring the smallest clusters in terms of
population, that is clusters $4$ and $5$, clusters $3$, $7$ and $10$ also have
relatively high mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
low standard deviations relative to the mean speeds, suggesting reliability and
resilience, as well as quality broadband services. Cluster $7$ has a higher ratio of
standard deviation to mean, but there is less difference in average speeds between
the morning and evening peaks than in clusters $3$ and $10$.

@@

Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
metropolitan area, even though the four local authorities of cluster $7$ include
no central urban boroughs and only one LAD that is part of a metropolitan area of
governance - Tameside in Greater Manchester. This may explain why those in cluster
$7$ are the second least likely to be served by Virgin Media.
It is also a demonstration of the complexity of both experienced broadband 
upload speeds as captured by time-series clustering, and the geography of 
first level digital divides as a product of quality as well as availability.
In comparison, cluster $1$, which our analysis suggests lacks broadband resilience, 
contains five, mainly rural authorities, but they are closer to a metropolitan area 
than authorities in cluster $3$, although furthest from London, perhaps because 
they are scattered around the country.

@@

Yet we need to consider the results for other variables in Table \ref{aux} to better
determine whether clusters $1$ and $13$, which appear to suffer most from poor
quality internet services, are also more likely to have a low skilled workforce,
less able to benefit from telecommuting.
Cluster $1$ has the lowest proportion of educated people,
and the lowest earnings, despite recording the highest proportion of managerial and
professional jobs, and the second highest proportion of tech jobs.
Rural areas such as those in cluster $1$ are home to many retired people,
which might explain these results or perhaps, as these figures are relative,
we could note that cluster $1$ also has a greater proportion of skilled trades
than other clusters. It seems likely that the first level digital divide reinforces
other inequities in cluster $1$, whilst those in cluster $13$ are more likely
to earn more and have a better education despite poor internet services.
Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
middle in terms of job density. If this is an indication of a high number of SMEs,
it could explain the variable internet quality, considering that small businesses
have not been seen as the most valuable customers for higher speed broadband services
if they are not located near residential customers [@ofcom2016].

Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
noise within the averages we use to measure internet quality and reliability.
Our results indicate average mean speeds, and Table \ref{aux} shows that
cluster $6$ also falls towards the middle of the clusters on many of the
socioeconomic variables. It ranks third or fourth out of the eleven in terms of
the likelihood of having a higher proportion of managerial, professional, and
tech jobs, as well as higher earnings, but is fourth from bottom for educational
attainment. The LADs in cluster $6$ are also diverse, with few truly remote areas,
but urban areas throughout England, including Birmingham, Leeds, Sheffield,
twelve London Boroughs, and many suburban areas and smaller cities like Oxford
and Cambridge. The capital cities of the other UK nations, Belfast, Cardiff and
Edinburgh, are also in this cluster, suggesting perhaps that the lower level of
reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand,
e.g. for telecommuting, when it is not supported by the most resilient internet
connections.

In contrast, Clusters $9$ and $11$ enjoy resilient internet connections, 
but LADs in Cluster $9$ are more likely to host highly educated individuals 
with higher earnings than cluster $11$, which has a negative coefficient for 
the NVQ4 variable. Cluster $11$ has fewer businesses per inhabitant and the 
second highest job density and cluster $9$ the second lowest.
These coefficients might indicate that resilient broadband infrastructure
generates higher returns for those in cluster $9$, where slightly more slowdown in
the AM peak was detected (see Table \ref{up.cluster.descr}).
LADs in cluster $9$ are also less likely to be in the South and those in cluster $11$
more likely, although the coefficient for cluster $9$ might be skewed by
the presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
different economic profile than England. Still, both clusters consist of larger
LADs in terms of population, including the city centre among four of the ten
districts of Greater Manchester, eight London Boroughs, four of the seven
constituent authorities of the West Midlands Combined Authority, Nottingham, and
both Portsmouth and Southampton. Among the $65$ LADs are also a number of other
tightly bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
Stevenage, and urban areas at the centre of less confined districts like Burnley,
the digital hub of Milton Keynes, and Northampton.
These urban locations are on the right side of the first level digital divide,
and the regression results suggest that internet resilience is supporting a wide
range of small and large urban economies.

LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level
digital divide, experiencing higher than average mean speeds and ranking high to
middle on measures of service reliability.
However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
workers to take advantage of working from home opportunities.
Cluster $2$ is comprised of just two LADs, with the lowest job density of any
cluster. Cluster $10$ is comprised of four LADs, including two
peripheral suburban areas of Birmingham, the East London Borough of Newham,
and Dundee. Suburbs are considered the most likely urban form in which telecommuters
live [@e2018does], and Dundee has a reputation for tech startups [@technation2017].
LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
district of Greater Manchester, the villages between York and Leeds, and a couple of
small West Midlands towns have the highest earnings, and thus may also be making the
most of their resilient internet. In contrast, LADs in cluster $3$ are associated
more with lower skills and have the highest proportion of individuals in machine
operation jobs. Arguably, LADs in clusters $3$ benefit the least from their
resilient internet infrastructure in an era when working from home became a vehicle
for economic resilience.

Finally, clusters $8$ and $12$ consist of LADs enjoying close to average
mean upload speeds, but opposing patterns of internet reliability during the morning
peak. LADs in cluster $8$ are likely to host more individuals employed in tech and
administrative occupations than any other cluster, as well as many in managerial
occupations, whilst the opposite applies to skilled trade and leisure occupations.
These LADs are characterised by a very small positive likelihood of more educated
residents, but are not significantly likely to earn more than other clusters.
Including suburbs near Leicester, Cardiff, Birmingham, the green belt area south of
London straddling the M25 motorway, and rural West Devon, the LADs in cluster $8$
seem to likely to have the skills and occupations that would benefit from
quality internet services, but suffer from poor internet resilience and reliability.

In comparison, LADs in cluster $12$ are less likely to be able to benefit from
quality internet services if they had them, with fewer individuals achieving
NVQ4 or better and lower levels of occupations that would benefit from homeworking.
Yet LADs in this cluster benefit from the second highest level of earnings, and
have the largest stock of businesses per inhabitant and the highest job density.
This density of businesses could be why cluster $12$ is the only cluster with higher
speeds in the morning peak during the study period. If people are at home, business
premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
spread across twelve LADs, including the City of Westminster, 
another London Borough, and two others outside London, but also a LAD in
Northern Ireland, two in Wales, Preston in the Northwest, and Fenland in the East of
England. This spatial diversity demonstrates that the temporal clustering of internet
resilience is not necessarily spatially dependent, and digital divides do not
necessarily overlap with economic ones.

@@

This paper offers a new perspective on telecommuting from the viewpoint of the
complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods,
a family of unsupervised machine learning techniques which are rarely utilised
in geographical research. Fast, reliable internet connections are
necessary for the population to be able to work from home.
Although not every place hosts individuals in occupations which allow for
telecommuting nor with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that without good internet connectivity,
these places would struggle more to achieve economic resilience in a period like
the current pandemic in any case.
Indeed, our analysis demonstrated that the temporal profiles of twelve of our
thirteen clusters had slower upload speeds in the morning than in the evening.
The opposite is likely to have been the norm prior to the pandemic, as the level of
demand and bandwidth management is the most common cause of temporal variation in
experienced speeds, and why evening download speeds, rather than daytime upload
speeds, have been used to benchmark the performance of internet services.
Thus, the new patterns can be taken as evidence of widespread telecommuting
and other daytime internet use which changed the temporal profile of
internet activity throughout the UK, not just in areas with more digital industry
or better skills.

Upload speeds have not previously been seen as integral to universal service,
considering there has never before been such extreme demand for telecommuting
and operations such as video calls.
This may be why the average upload speeds for the largest two clusters
-- $6$ and $13$ -- contain so many LADs that are centres of the knowledge economy,
from Oxford and Cambridge in the former, to areas like Reading in the latter, 
which has the highest concentration of digital businesses in the country 
[@technation2017].
Thus, whilst some areas suffer from an intersection of digital and economic divides,
such as those in cluster $1$, others found their digital infrastructure to be less
reliable when confronted with the sudden change in the timing and type of demand.
Still other LADs, such as Milton Keynes in cluster $9$ were able to benefit both
from reliable internet connections and populations which were familiar with working
from home and could capitalise on their digital infrastructure.
Yet the nuanced picture we gained through our analysis of the UK case study suggests
that quality internet connectivity would also enable other LADs to gain ground
during the pandemic despite lower skills or earnings currently, especially in a
future where telecommuting might be a more common means of accessing work
and broadband services must be fit for purpose.

<!-- WFH broader consequences-->
Indeed, the long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span various aspects of economy and society: from changes to
transportation planning due to altered commuting patterns to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713; @ELLDER2020102777], and from productivity
and innovation changes to changes in agglomeration externalities and the
attraction of large cities [@econobs].
Further research may be able to measure the economic resilience of the different
clusters of places discussed in this paper once this pandemic is firmly past.
However, our analysis demonstrates that the economic resilience made possible by
working from home cannot be understood without considering the underpinning digital
divides and cannot be achieved without planning for how the levels of digital,
social and economic divides might intersect.

>>>>

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds
during the spring 2020 lockdown,
may reinforce or redress the spatial and social dimensions of digital divisions.
We employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of education, occupation and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era.

@@

As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and interactions
between digital divisions geographically and socioeconomically.
By asking how resilient broadband speeds, and particularly upload speeds, are as
experienced in different parts of the UK during a time of extreme demand,
we interrogate which places benefit from the greater economic resilience
digital technologies can offer.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand their structural and spatial development pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across clusters of UK local authorities and then assesses whether these clusters
replicate or repudiate other socio-economic and geographic patterns of economic
resilience.
We conclude with a discussion of the insights we have gained from our new
perspective on digital divisions.

@@

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies.
Rural areas, predictably, fare worse, yet as @blank2018local highlight, variation
in individual internet uptake and use is a product of more complex spatial and 
demographic characteristics than simple rurality or urbanisation.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been investigated, in part because telecommuting has
not previously been a cause of concentrated demand and pressure on internet services.

@@

The extreme demand during the pandemic thus provides a new opportunity to understand
how internet infrastructure quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive 
video conferencing required in order to avoid the face-to-face contact that could 
increase the spread of infection.
We seek to answer how internet service resilience could contributes to or reduce 
economic resilience when the latter is dependent upon the capability to 
work from home.
We also aim to improve our understanding of the impact of first level digital
divisions on telecommuting, and whether this results in more fundamental third
level digital divisions than has previously been perceived.

@@

![Speed tests over time, 2019 \label{test2019}](figures/time.var.plot2019.png){height=50% width=85%}

@@

Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
clusters $1$, $13$, and $6$. However, mean upload speeds are much lower between
$9:00-10:59$ than between $19:00-20:59$ in cluster $8$, but slightly faster in the
morning than in the evening in cluster $12$.
Indeed, cluster $12$ is the only cluster to experience higher speeds in the morning
peak compared to the evening, and thus the only cluster where the temporal profile
of internet use is closer to what might have been expected pre-pandemic.
Among the other clusters, however, the reliability of internet services
during the working day still varies considerably.
Interpreting this variation from the large spikes and dips shown on
Figures \ref{UpClusterL} and \ref{UpClusterS} is difficult,
but the statistics in Table \ref{up.cluster.descr} show that clusters $9$ and $11$
have the most reliable internet services.
The ratio of standard deviation to mean in both these clusters is lowest,
and speeds are only about 2% slower in the morning than the evening.
Mean speeds are also higher than in any other cluster, excluding cluster $2$,
where measures of reliability suggest poorer performance.

Thus, broadband services in clusters $9$ and $11$, 
home to over twelve million people, performed the best during the study period, 
in terms of both quality and reliability.
In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
with the lowest points occurring, on average, between $6:00-7:00$ on Monday morning, 
$14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday, not the workday peak. 
Furthermore, these slow periods offer better speeds than the average hourly
profile of cluster $13$. Clusters $3$, $7$ and $10$ also have relatively high 
mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
low standard deviations relative to the mean speeds, suggesting reliability and
resilience, as well as quality broadband services. Cluster $7$ has a higher ratio of
standard deviation to mean, but there is less difference in average speeds between
the morning and evening peaks than in clusters $3$ and $10$.

@@

Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
metropolitan area, even though the four local authorities of cluster $7$ include
no central urban boroughs and only one LAD that is part of a metropolitan area of
governance - Tameside in Greater Manchester. This may explain why those in cluster
$7$ are the second least likely to be served by Virgin Media.
It is also a demonstration of the complexity of both experienced broadband 
upload speeds as captured by time-series clustering, and the geography of 
first level digital divides as a product of quality as well as availability.
In comparison, cluster $1$, which our analysis suggests lacks broadband resilience, 
contains five, mainly rural authorities, but they are closer to a metropolitan area 
than authorities in cluster $3$, although furthest from London, perhaps because 
they are scattered around the country -- see Figure \ref{map.up.clusters}.

@@

Yet we need to consider the results for other variables in Table \ref{aux} to better
determine whether clusters $1$ and $13$, which appear to suffer most from poor
quality internet services, are also more likely to have a low skilled workforce,
less able to benefit from telecommuting.
Cluster $1$ has the lowest proportion of educated people,
and the lowest earnings, despite recording the highest proportion of managerial and
professional jobs, and the second highest proportion of tech jobs.
Rural areas such as those in cluster $1$ are home to many older, retired people 
[@blank2018local], which might explain these results or perhaps, 
as these figures are relative, we could note that cluster $1$ also has a greater 
proportion of skilled trades than other clusters. 
In any case, it appears that the first level digital divide 
reinforces other inequities in cluster $1$, where the even slower than average 
morning upload speeds shown in Table \ref{up.cluster.descr} suggest that internet
users were more active during the working day. Meanwhile, those in cluster $13$ are 
more likely to earn more and have a better education despite poor internet services.
Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
middle in terms of job density. If this is an indication of a high number of SMEs,
it could explain the variable internet quality, considering that small businesses
have not been seen as the most valuable customers for higher speed broadband services
if they are not located near residential customers [@ofcom2016].

Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
noise within the averages we use to measure internet quality and reliability.
Our results indicate average mean speeds, and Table \ref{aux} shows that
cluster $6$ also falls towards the middle of the clusters on many of the
socioeconomic variables. It ranks third or fourth out of the eleven in terms of
the likelihood of having a higher proportion of managerial, professional, and
tech jobs, as well as higher earnings, but is fourth from bottom for educational
attainment. The LADs in cluster $6$ are also diverse, with few truly remote areas,
but urban areas throughout England, including Birmingham, Leeds, Sheffield,
twelve London Boroughs, and many suburban areas and smaller cities like Oxford
and Cambridge. The capital cities of the other UK nations, Belfast, Cardiff and
Edinburgh, are also in this cluster, suggesting perhaps that the lower level of
reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand,
e.g. for telecommuting, despite lacking the most resilient internet connections.

In contrast, Clusters $9$ and $11$ enjoy resilient internet connections, 
but LADs in Cluster $9$ are more likely to host highly educated individuals 
with higher earnings than cluster $11$, which has a negative coefficient for 
the NVQ4 variable. Cluster $11$ has fewer businesses per inhabitant and the 
second highest job density and cluster $9$ the second lowest.
These coefficients might indicate that resilient broadband infrastructure
generates higher returns for those in cluster $9$, where slightly more slowdown in
the AM peak was detected -- see Table \ref{up.cluster.descr}.
LADs in cluster $9$ are less likely to be in the South and those in cluster $11$
more likely, although the coefficient for cluster $9$ might be skewed by
the presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
different economic profile than England. Still, both clusters consist of larger
LADs in terms of population, including districts within five of England's ten
largest cities, and a number of other stand-alone urban areas and large market towns 
-- see Appendix [1](#appendix1).
These urban locations are on the right side of the first level digital divide,
and the regression results suggest that internet resilience is supporting a wide
range of small and large urban economies.

LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level
digital divide, experiencing higher than average mean upload speeds and ranking 
high to middle on measures of service reliability.
LADs in clusters $2$ and $10$ are also more likely to have highly skilled
workers to take advantage of working from home opportunities.
Cluster $2$ is comprised of just two LADs, with the lowest job density of any
cluster. Cluster $10$ is comprised of four LADs, including two
peripheral suburban areas of Birmingham, the East London Borough of Newham,
and Dundee. Suburbs are considered the most likely urban form in which telecommuters
live [@e2018does], and Dundee has a reputation for tech startups [@technation2017].
LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
district of Greater Manchester, and some other Midlands and North Yorkshire towns 
and villages have the highest earnings, and thus may also be making the
most of their resilient internet. In contrast, LADs in cluster $3$ are associated
more with lower skills and have the highest proportion of individuals in machine
operation jobs. Arguably, LADs in clusters $3$ benefit the least from their
resilient internet infrastructure in an era when working from home became a vehicle
for economic resilience.

Finally, clusters $8$ and $12$ consist of LADs enjoying close to average
mean upload speeds, but opposing patterns of internet reliability during the morning
peak. LADs in cluster $8$ are likely to host more individuals employed in tech and
administrative occupations than any other cluster, as well as many in managerial
occupations, whilst the opposite applies to skilled trade and leisure occupations.
These LADs are characterised by a very small positive likelihood of more educated
residents, but are not significantly likely to earn more than other clusters.
Including suburbs near Leicester, Cardiff, and Birmingham and south of London, 
the LADs in cluster $8$ seem to likely to have the skills and occupations that 
would benefit from quality internet services, but suffer from poor internet 
resilience and reliability.

In comparison, LADs in cluster $12$ are less likely to be able to benefit from
quality internet services if they had them, with fewer individuals achieving
NVQ4 or better and lower levels of occupations that would benefit from homeworking.
Yet LADs in this cluster benefit from the second highest level of earnings, and
have the largest stock of businesses per inhabitant and the highest job density.
This density of businesses could be why cluster $12$ is the only cluster with higher
speeds in the morning peak during the study period. If people are at home, business
premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
spread across twelve LADs, from London to Wales as listed in Appendix [1](#appendix1). 
This spatial diversity demonstrates that the temporal clustering of internet
resilience is not necessarily spatially dependent, and digital divides do not
necessarily overlap with economic ones.

@@

This paper offers a new perspective on telecommuting from the viewpoint of the
complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods,
a family of unsupervised machine learning techniques which are rarely utilised
in geographical research. Fast, reliable internet connections are
necessary for the population to be able to work from home.
Although not every place hosts individuals in occupations which allow for
telecommuting nor with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that places without good internet 
connectivity will still struggle more than other places to achieve economic 
resilience in a period like the current pandemic when internet resilience is 
so vital.
Indeed, our analysis demonstrated that the temporal profiles of twelve of our
thirteen clusters had slower upload speeds in the morning than in the evening.
The opposite is likely to have been the norm prior to the pandemic, as level of
demand and bandwidth management is the most common cause of temporal variation in
experienced speeds, and why evening download speeds, rather than daytime upload
speeds, have been used to benchmark the performance of internet services.
Thus, the new patterns can be taken as evidence of widespread telecommuting
and other daytime internet use which changed the temporal profile of
internet activity throughout the UK, not just in areas with more digital industry
or better skills.

Upload speeds have not previously been seen as integral to universal service,
considering there has never before been such extreme demand for telecommuting
and operations such as video calls.
This may be why the average upload speeds for the largest two clusters
-- $6$ and $13$ -- contain so many LADs that are centres of the knowledge economy,
from Oxford and Cambridge in the former, to areas like Reading in the latter, 
which has the highest concentration of digital businesses in the country 
[@technation2017].
Thus, whilst some areas suffer from an intersection of digital and economic divides,
such as those in cluster $1$, other areas, including cluster $8$ as well as $6$ and
$13$ also found their digital infrastructure to be less than reliable when 
confronted with the sudden change in the timing and type of demand.
In contrast, LADs such as the digital hub, Milton Keynes, in cluster $9$ were able 
to benefit both from reliable internet connections and populations which were 
familiar with working from home and could capitalise on their digital infrastructure.
Yet the nuanced picture we gained through our analysis of the UK case study suggests
that quality internet connectivity may also have enabled other LADs, such as those in
cluster $3$ or $11$ to gain ground during the pandemic despite lower skill levels.

<!-- WFH broader consequences-->
Digital infrastructure which considers upload speeds and working day reliability as 
well as availability are likely to be particularly important in a future where
telecommuting might be a more common means of accessing work and broadband services 
must be fit for purpose, although the long-term effects of such drastic changes in 
telecommuting and attitudes towards working from home are difficult to predict. 
Nevertheless, they span various aspects of economy and society: from changes to
transportation planning due to altered commuting patterns to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713; @ELLDER2020102777], and from productivity
and innovation changes to changes in agglomeration externalities and the
attraction of large cities [@econobs].
Further research may be able to measure the economic resilience of the different
clusters of places discussed in this paper once this pandemic is firmly past.
However, our analysis demonstrates that the economic resilience made possible by
working from home cannot be understood without considering the underpinning digital
divides and cannot be achieved without planning for how the levels of digital,
social and economic divides might intersect.
